Naturalness and Long-Lived Beyond the Standard Model Particles

### 2017-08-07 tex/draft.tex
Integrate jamie, reynolds comments and other little fixes

    The fermions and gauge bosons can be classified according to the forces with which they interact. The fermions are further divided into six quarks, which carry color and interact via the strong force, and six leptons, which do not. Furthermore, all six quarks and three of the leptons (the electron, muon, and tau) carry electric charge and interact electromagnetically. All fermions interact via the weak force. The gauge bosons include the photon, which communicates the electromagnetic force; the $\mathrm{W}^+$, $\mathrm{W}^-$, and $\mathrm{Z}$, which communicate the weak force; and eight gluons that communicate the strong force. Of these, only the $\mathrm{W}^+$ and $\mathrm{W}^+$ are electrically charged and only the gluons carry color charge.

@@

    SM parameters, such as masses and couplings, are corrected by interactions with virtual particles through loop diagrams. Fig.~\ref{loop_diagrams}, for example, shows interactions that correct the Higgs and electron masses. Because the Higgs is a fundamental scalar, its corrections depend quadratically on the cutoff scale, $\Lambda$, which is the largest energy scale for which the SM is valid. The fermions and gauge bosons, on the other hand, only depend logarithmically on $\Lambda$~\cite{dine_naturalness}. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the bare Higgs mass and the terms that correct it.

@@

    \noindent \begin{figure}[htbp] \begin{center}
        \feynmandiagram [inline=(d.base), layered layout, medium, horizontal=b to c] {
            a [particle=\(H\)] -- [scalar] b,
            b -- [fermion, half left, edge label=\(t\)] c,
            c -- [fermion, half left] b,
            c -- [scalar] d [particle=\(H\)],
    };
    \quad + \quad
        \feynmandiagram [inline=(d.base), horizontal=a to b, layered layout, medium] {
            a [particle=\(H\)] -- [scalar] b [dot] 
            -- [out=135, in=45, loop, min distance=2.5cm, charged scalar, edge label=\(\tilde{t}\)] b 
            -- [scalar] c [particle=\(H\)],
        };
        \quad = \quad 0 \quad
        \caption{Corrections to the Higgs mass from the top quark (left) and stop squark (right) cancel in exact SUSY. The top and stop contributions are enhanced by the large top Yukawa coupling. Diagrams created with~\cite{tikz}.}
        \label{exact_susy_correction}
    \end{center} \end{figure}

@@

    The mechanism and scale of SUSY breaking is an open question with many competing answers that achieve different theoretical goals and lead to different TeV-scale phenomena. A handful of models with particular approaches to SUSY breaking that are particularly prominent or particularly relevant to naturalness or long-lived particles are gauge-mediated SUSY breaking (GMSB), anomaly-mediated SUSY breaking (AMSB), Split SUSY, and R-parity violating (RPV) SUSY.

\section{Long-Lived Particles}
    The SUSY models most relevant to this discussion are those that are motivated by naturalness and produce long-lived particles (LLPs), a signal that could be missed by common SUSY search techniques. LLPs are defined as particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point. This category includes everything from particles that decay before reaching the inner tracker to particles that propagate through the entire detector.

    In the SM and in BSM theories, LLPs result from any of the following conditions: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly virtual intermediate states, or (4) limited phase space in which to decay. In the SM, charge and baryon number conservation guarantee the absolute stability of the electron and proton as the lightest particles carrying their respective conserved quantities. Neutrons and muons, whose decays are shown in Fig.\ref{sm_llp_processes}, are long-lived for reasons (2), (3), and (4). Being pure weak decays, they have small coupling constants and decay through a W, which is far more massive than either parent particle. Finally, the proton and electron masses leave them little phase space in which to decay.

@@

                \diagram* {
                    (n1a) -- [fermion] (n1b) -- [fermion] (p1),
                    (n2) -- [fermion] (p2),
                    (n3) -- [fermion] (p3),
                    (n1b) -- [boson, edge label=\(\W^{-}\)] (w),
                    (w) -- [fermion] (lc),
                    (ln) -- [fermion] (w),
                };
                \draw [decoration={brace}, decorate] (n3.south west) -- (n1a.north west)
                node [pos=0.5, left] {\(n\)};
                \draw [decoration={brace}, decorate] (p1.north east) -- (p3.south east)
                node [pos=0.5, right] {\(p\)};
            \end{feynman}
            \end{tikzpicture}
            \qquad\qquad
            \feynmandiagram [layered layout, horizontal=a to b] {
                a [particle=\(\mu\)] -- [fermion] b,
                b -- [fermion] d [particle=\(\nu_{\mu}\)],
                b -- [boson, edge label=\(\W^{-}\)] c,
                c -- [fermion] e [particle=\(e\)],
                c -- [anti fermion] f [particle=\(\nu_{e}\)],
            };
            \caption{Long-lived decays of the neutron and muon. Diagrams created with~\cite{tikz}.}
            \label{sm_llp_processes}
        \end{center} \end{figure}

@@

\section{The CMS Detector}
    \subsection{Overview}
        The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from proton-proton collisions at the Large Hadron Collider (LHC), a circular collider that ran with center-of-mass energy \SI{7}{\tera\electronvolt} in 2010-11, \SI{8}{\tera\electronvolt} in 2012, and \SI{13}{\tera\electronvolt} from 2015 through to the present. To reconstuct final-state particles, CMS uses a particle-flow algorithm that combines information from several subdetectors nested radially in and around a \SI{3.8}{\tesla} superconducting solenoid. All together, CMS is \SI{21.6}{\m} long, \SI{14.6}{\m} across, and weighs approximately \SI{12500}{t}~\cite{cms_experiment}. The subdetectors that make up CMS are shown in Fig.~\ref{cms} and described below.

@@

    \subsection{Tracker}
        In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions. Inside \SI{20}{cm}, the pixel tracker uses small silicon pixels to keep the occupancy near \SI{0.1}{\percent} despite the high particle flux. Outside \SI{20}{\cm}, the silicon strip tracker (SST) achieves similar occupancy with larger silicon strips~\cite{cms_tdr}.\fxnote{check occupancy numbers}

@@

    \subsection{Electromagnetic Calorimeter}
        After traversing the inner tracker, particles next encounter the electromagnetic calorimeter (ECAL). As a homogeneous scintillation calorimeter, ECAL uses \num{61200} lead tungstate crystals in the barrel and \num{7324} in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (\SI{80}{\percent} of light within \SI{25}{\nano\s}), compact (radiation length = \SI{0.89}{cm}), fine-grained (Moli\`ere radius = \SI{2.2}{cm}), and radiation hard (up to 10 Mrad) calorimeter. The main drawback is the relatively low light yield (\SI{30}{photon\per\mega\electronvolt}), which necessitates photodetectors with intrinsic gain that work in magnetic fields\cite{cms_experiment, cms_tdr}. Energy resolution is varies from approximately \num{1}-\SI{5}{\percent} depending on the amount of material traversed before reaching ECAL~\cite{ecal_performance}.

@@

        The barrel section (HB) spans a pseudorapidity range of \num{-1.4} to \num{1.4} and has \num{2304} towers \fxnote{check}, each of which contains \num{14} \fxnote{check} brass plates and covers an area of \num{0.087} x \SI{0.087}{in} $\eta-\phi$. In addition, an extra layer (or two at $\eta = \num{0}$) of scintillator tiles sits just outside the solenoid. This extra layer, known as hadron outer (HO), spans a pseudorapidity range of \num{-1.26} to \num{1.26} and increases the minimum effective HCAL interaction length to greater than \num{11.8}.

@@

    \subsection{Solenoid}
        CMS employs a \SI{4}{T} superconducting solenoid to aid in the measurement of charged particle momenta. The solenoid consists of \num{2168} turns of Nb-Ti superconducting cable and has an inner diameter of \SI{6}{m} and a length of \SI{12.5}{m}. The flux returns through an iron yoke that also houses the muon system. The significant  bending power helps CMS meet its physics goals of unambiguous sign determination and momentum resolution of \SI{10}{\percent} for muons up to \SI{1}{\tera\electronvolt}\cite{cms_experiment}.

@@

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.6\textwidth]{images/muon_pt_resolution}
        \caption{Transverse momentum resolution in the tracker and muon system for small (left) and large (right) pseudorapidity~\cite{cms_image}.}
        \label{muon_pt_resolution}
        \end{center} \end{figure}

@@

\section{LHC Searches for Long-lived Particles} \label {analyses}
    \fxnote{mention which data was used for each search} 
    \subsection{Overview}
        LLPs leave distinct signatures that differentiate them from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were mainly designed to reconstruct stable particles from prompt decays, so LLP searches are often forced to stray from the traditional analysis path in terms of object reconstruction and background estimation. This report focuses on four complementary LLP signatures: displaced vertices, disappearing tracks, low-velocity particles, and particles that come to a stop before decaying. See Fig.~\ref{jamie_cartoon} for a cartoon of several possible LLP signatures.

@@

    \subsection{Displaced Vertices}
        LLPs that decay before reaching the inner tracker can be identified with a displaced vertex (DV). ATLAS performs a search for such particles by looking for at least one DV along with large MET and large track multiplicity~\cite{atlas_displaced}. The search is sensitive to LLPs with lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range.

@@

        \noindent \begin{figure}[htbp] \begin{center}
        \feynmandiagram [layered layout, horizontal=a to b, small] {
            a [particle=\(\tilde{g}\)] -- [plain, gluon] b,
            b -- [anti charged scalar, edge label=\(\tilde{q}\)] c,
            b -- [fermion] d [particle=\(q\)],
            c -- [anti fermion] e [particle=\(q\)],
            c -- [plain, boson] f [particle=\(\tilde{\chi}_1^0\)],
        };
        \caption{Long-lived gluino decay through highly virtual squark. Diagram created with~\cite{tikz}.}
        \label{displaced_process}
        \end{center} \end{figure}

@@

        The authors minimize the effect of hadronic interactions by vetoing the material-dense regions of the detector (see Fig.~\ref{material_map}) and estimate the residual background by extrapolating from the low-mass region, which is dominated by hadronic interactions. After a dedicated study, the background from merged vertices was found to be negligible. Finally, the background from crossing tracks is determined from a model that adds pseudo-tracks to data vertices.

@@

        After estimating $\num{0.02} \pm \num{0.02}$ background events, exactly zero are observed. The result is then interpreted as an upper limit on the gluino mass and pair-production cross section as a function of gluino lifetime, shown in Fig.~\ref{displaced_limit}. For an LSP neutralino mass of \SI{100}{\giga\electronvolt}, the gluino mass limit is approximately \SI{2000}{\giga\electronvolt} for lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range.

@@

    \subsection{Disappearing Tracks}
        Certain models with nearly mass-degenerate LSP and NLSP have the potential to present a remarkable signal: charged particles that seem to disappear in the tracker. Specifically, models with winos as the lightest gauginos, such as many AMSB models, will have a chargino NLSP with an observable lifetime (\SI{0.2}{\nano\s} is a typical estimate that eventually decays to the neutralino LSP and, commonly, a pion. The nearly degenerate masses have two important phenomenological roles: (1) increase the chargino lifetime, and (2) ensure low-momentum decay products. The low-momentum pion is then unlikely to be reconstructed, which results in a disappearing track~\cite{atlas_disappearing}.

        Both ATLAS and CMS looked for disappearing tracks during Run I, but the tracker improvements installed in ATLAS before 2015 datataking and CMS before 2017 datataking should increase sensitivity to decays at shorter radii. The ATLAS search presented here is sensitive to LLP lifetimes of approximately \SI{10}{\pico\s} to \SI{10}{\nano\s}.

@@

        Disappearing track candidates, called pixel tracklets, are selected with a looser track reconstruction algorithm that runs over hits that were not associated to tracks by the standard reconstruction algorithm. The pixel tracklets are required to pass several isolation, transverse momentum ($\pT$), and quality requirements such as number of pixel layer hits. Finally, the disappearing tracks are selected by requiring exactly zero associated hits in the subsequent tracking layer. Event selection requires one pixel tracklet, no electron or muon candidates (to suppress $\ttbar$ and $\W/\Z + \mathrm{jets}$ processes, and the $\MET$ and jet requirements associated with either of the two production modes described above.

        Pixel tracklets can be faked by random combinations of unrelated hits from nearby particles or by tracks whose pixel hits and next tracking layer hits are not properly associated, either due to a hadron suffering a hard scatter or a lepton radiating a hard photon. These backgrounds come primarily from $\ttbar$ and $\W/\Z + \mathrm{jets}$ processes, and estimates for each type of background are obtained from data.

@@

    \subsection{Stopped Particles}
        \fxnote{Add calorimeter search?}  
        \fxnote{too redundant with overview?}
        With long enough lifetimes and sufficient mass, LLPs could come to rest in the detector before decaying. The eventual decay would then lead to energy deposits in the calorimeters or hits in the muon system that could be out of time with proton bunch crossings. A recent CMS analysis looked for such a signal in dimuon final states using the full 2015 and 2016 datasets~\cite{cms_stopped}.

@@

        After estimating the background by extrapolating data from the cosmic-enriched region to the signal region and determining systematic uncertainties related to the MC simulation, trigger acceptance, and luminosity, the authors calculate the expected background for LLP lifetimes ranging from $10^{-7}$ to $10^6$ s. In all cases, the expected background is less then one event. Exactly zero events are observed in every lifetime scenario. As shown in Fig.~\ref{stopped_limits}, these results are then interpreted as upper limits on the signal production cross section and mass\fxnote{mass of what?}, which are the first limits for stopped particles that decay to muons at the LHC.

@@

        HSCPs will often be misidentified or missed entirely by standard reconstruction algorithms and analysis criteria, which tend to assume $\frac{v}{c} \approx \num{1}$ and $\lvert q \rvert = e$. Inverting these assumptions provides two handles for a dedicated HSCP search in the context of three benchmark models: (1) HSCPs that form R-hadrons that either change the sign of their electric charge or become neutral before reaching the MS, (2) lepton-like HSCPs (quasi-stable sleptons) that are produced either through decays of squarks or gluinos or directly if squarks and gluinos are kinematically inaccessible, and (3) long-lived lepton-like fermions with arbitrary electric charge. This discussion will focus on the first two models, as they bear more directly on naturalness and SUSY.

@@

        The authors estimate the signal-region background using an ABCD method in which events are binned in a 2-D plane with uncorrelated axes and the ratio of events in two control regions is used to extrapolate into the signal region from a third control region. Systematic uncertainties are introduced in background estimation, signal acceptance, and integrated luminosity.

@@

    \noindent \begin{figure}[htbp] \begin{center}
    \includegraphics[width=.8\textwidth]{images/atlas_susy_summary}
    \caption{Summary of current experimental limits on SUSY models from ATLAS searches.}
    \label{atlas_susy_summary}
    \end{center} \end{figure}

@@

        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.3\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/cornering_effective_100}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.3\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/cornering_rpv_100}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.3\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/cornering_rpv_effective_100}
        \end{center} \end{subfigure}
            \caption{Comparing experimental limits to $\Delta < 10$ naturalness bounds for Effective (left), RPV (center), and Effective+RPV (right) SUSY models ~\cite{cornering}.}
        \label{cornering_limits}
        \end{center} \end{figure}

>>>>

    The fermions and gauge bosons can be classified according to the forces with which they interact. The fermions are further divided into six quarks, which carry color and interact via the strong force, and six leptons, which do not. Furthermore, all six quarks and three of the leptons (the electron, muon, and tau) carry electric charge and interact electromagnetically. All fermions interact via the weak force. The gauge bosons include the photon, which communicates the electromagnetic force; the $\mathrm{W}^+$, $\mathrm{W}^-$, and $\mathrm{Z}$, which communicate the weak force; and eight gluons that communicate the strong force. Of these, only the $\mathrm{W}^+$ and $\mathrm{W}^-$ are electrically charged and only the gluons carry color charge.

@@

    SM parameters, such as masses and couplings, are corrected by interactions with virtual particles through loop diagrams. Fig.~\ref{loop_diagrams}, for example, shows interactions that correct the Higgs and fermion masses. Because the Higgs is a fundamental scalar, its corrections depend quadratically on the cutoff scale, $\Lambda$, which is the largest energy scale for which the SM is valid. The fermions and gauge bosons, on the other hand, only depend logarithmically on $\Lambda$~\cite{dine_naturalness}. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the bare Higgs mass and the terms that correct it.

@@

    \noindent \begin{figure}[htbp] \begin{center}
        \feynmandiagram [inline=(d.base), layered layout, medium, horizontal=b to c] {
            a [particle=\(H\)] -- [scalar] b,
            b -- [fermion, half left, edge label=\(t\)] c,
            c -- [fermion, half left] b,
            c -- [scalar] d [particle=\(H\)],
    };
    \quad + \quad
        \feynmandiagram [inline=(d.base), horizontal=a to b, layered layout, medium] {
            a [particle=\(H\)] -- [scalar] b [dot] 
            -- [out=135, in=45, loop, min distance=2.5cm, charged scalar, edge label=\(\tilde{t}\)] b 
            -- [scalar] c [particle=\(H\)],
        };
        \quad = \quad 0 \quad
        \caption{Corrections to the Higgs mass from the top quark (left) and stop squark (right) cancel in exact SUSY. The top and stop contributions are enhanced by the large top Yukawa coupling.}
        \label{exact_susy_correction}
    \end{center} \end{figure}

@@

    The mechanism and scale of SUSY breaking is an open question with many competing answers that achieve different theoretical goals and lead to different TeV-scale phenomena. A handful of models with particular approaches to SUSY breaking that are particularly prominent or particularly relevant to naturalness or long-lived particles are gauge-mediated SUSY breaking (GMSB), anomaly-mediated SUSY breaking (AMSB), Split SUSY, and R-parity violating (RPV) SUSY\fxnote{find a way to water this down or kill it}.

\section{Long-Lived Particles}
    The SUSY models most relevant to this discussion are those that are motivated by naturalness and produce long-lived particles (LLPs), a signal that could be missed by common SUSY search techniques. LLPs are defined as particles whose lifetimes are such that they decay a measurable distance from the nominal interaction point. This category includes everything from particles that decay before reaching the inner tracker to particles that propagate through the entire detector.

    In the SM and in BSM theories, LLPs can result from any of the following conditions: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly virtual intermediate states, or (4) limited phase space in which to decay. In the SM, charge and baryon number conservation guarantee the absolute stability of the electron and proton as the lightest particles carrying their respective conserved quantities. Neutrons and muons, whose decays are shown in Fig.\ref{sm_llp_processes}, are long-lived for reasons (2), (3), and (4). Being pure weak decays, they have small coupling constants and decay through a W, which is far more massive than either parent particle. Finally, the proton and electron masses leave them little phase space in which to decay.

@@

                \diagram* {
                    (n1a) -- [fermion] (n1b) -- [fermion] (p1),
                    (n2) -- [fermion] (p2),
                    (n3) -- [fermion] (p3),
                    (n1b) -- [boson, edge label=\(\W^{-}\)] (w),
                    (w) -- [fermion] (lc),
                    (ln) -- [fermion] (w),
                };
                \draw [decoration={brace}, decorate] (n3.south west) -- (n1a.north west)
                node [pos=0.5, left] {\(n\)};
                \draw [decoration={brace}, decorate] (p1.north east) -- (p3.south east)
                node [pos=0.5, right] {\(p\)};
            \end{feynman}
            \end{tikzpicture}
            \qquad\qquad
            \feynmandiagram [layered layout, horizontal=a to b] {
                a [particle=\(\mu\)] -- [fermion] b,
                b -- [fermion] d [particle=\(\nu_{\mu}\)],
                b -- [boson, edge label=\(\W^{-}\)] c,
                c -- [fermion] e [particle=\(e\)],
                c -- [anti fermion] f [particle=\(\nu_{e}\)],
            };
            \caption{Long-lived decays of the neutron and muon.}
            \label{sm_llp_processes}
        \end{center} \end{figure}

@@

\section{The CMS Detector}
    \subsection{Overview}
        The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from proton-proton collisions at the Large Hadron Collider (LHC), a circular collider that ran with center-of-mass energy \SI{7}{\tera\electronvolt} in 2010-11, \SI{8}{\tera\electronvolt} in 2012, and \SI{13}{\tera\electronvolt} from 2015 through to the present. To reconstruct final-state particles, CMS uses a particle-flow algorithm that combines information from several subdetectors nested radially in and around a \SI{3.8}{\tesla} superconducting solenoid. All together, CMS is \SI{21.6}{\m} long, \SI{14.6}{\m} across, and weighs approximately \SI{12500}{t}~\cite{cms_experiment}. The subdetectors that make up CMS are shown in Fig.~\ref{cms} and described below.

@@

    \subsection{Tracker}
        In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions. Inside \SI{20}{cm}, the pixel tracker uses small silicon pixels to keep the occupancy near \SI{0.1}{\percent} despite the high particle flux. Outside \SI{20}{\cm}, the silicon strip tracker (SST) achieves an occupancy of $\approx \SI{2}{\percent}$ with larger silicon strips~\cite{cms_tdr}.

@@

    \subsection{Electromagnetic Calorimeter}
        After traversing the inner tracker, particles next encounter the electromagnetic calorimeter (ECAL). As a homogeneous scintillation calorimeter, ECAL uses \num{61200} lead tungstate crystals in the barrel and \num{7324} in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (\SI{80}{\percent} of light within \SI{25}{\nano\s}), compact (radiation length = \SI{0.89}{cm}), fine-grained (Moli\`ere radius = \SI{2.2}{cm}), and radiation hard (up to 10 Mrad) calorimeter. The main drawback is the relatively low light yield (\SI{30}{photon\per\mega\electronvolt}), which necessitates photodetectors with intrinsic gain that work in magnetic fields\cite{cms_experiment, cms_tdr}. Energy resolution varies from approximately \num{1}-\SI{5}{\percent} depending on the amount of material traversed before reaching ECAL~\cite{ecal_performance}.

@@

        The barrel section (HB) spans a pseudorapidity range of \num{-1.4} to \num{1.4} and is segmented into \num{32} towers in $\eta$ and \num{64} in $\phi$ that each contain 17 active scintillator layers. In addition, an extra layer (or two at $\eta = \num{0}$) of scintillator tiles sits just outside the solenoid. This extra layer, known as hadron outer (HO), spans a pseudorapidity range of \num{-1.26} to \num{1.26} and increases the minimum effective HCAL interaction length to greater than \num{11.8}.

@@

    \subsection{Solenoid}
        CMS employs a \SI{3.8}{T} superconducting solenoid to aid in the measurement of charged particle momenta. The solenoid consists of \num{2168} turns of Nb-Ti superconducting cable and has an inner diameter of \SI{6}{m} and a length of \SI{12.5}{m}. The flux returns through an iron yoke that also houses the muon system. The significant  bending power helps CMS meet its physics goals of unambiguous sign determination and momentum resolution of \SI{10}{\percent} for muons up to \SI{1}{\tera\electronvolt}\cite{cms_experiment}.

@@

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.65\textwidth]{images/muon_pt_resolution}
        \caption{Transverse momentum resolution in the tracker and muon system for small (left) and large (right) pseudorapidity~\cite{cms_image}.}
        \label{muon_pt_resolution}
        \end{center} \end{figure}

@@

\section{LHC Searches for Long-lived Particles} \label {analyses}
    \subsection{Overview}
        LLPs leave distinct signatures that differentiate them from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were mainly designed to reconstruct stable particles from prompt decays, so LLP searches are often forced to stray from the traditional analysis path in terms of object reconstruction and background estimation. This report focuses on four complementary LLP signatures: displaced vertices, disappearing tracks, low-velocity particles, and particles that come to a stop before decaying. See Fig.~\ref{jamie_cartoon} for a cartoon of several possible LLP signatures.

@@

    \subsection{Displaced Vertices}
        LLPs that decay before reaching the inner tracker can be identified with a displaced vertex (DV). ATLAS performs a search for such particles by looking for at least one DV along with large MET and large track multiplicity~\cite{atlas_displaced}. The search uses \SI{32.7}{\femto\barn^{-1}} of \SI{13}{TeV} data and is sensitive to LLPs with lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range.

@@

        \noindent \begin{figure}[htbp] \begin{center}
        \feynmandiagram [layered layout, horizontal=a to b, small] {
            a [particle=\(\tilde{g}\)] -- [plain, gluon] b,
            b -- [anti charged scalar, edge label=\(\tilde{q}\)] c,
            b -- [fermion] d [particle=\(q\)],
            c -- [anti fermion] e [particle=\(q\)],
            c -- [plain, boson] f [particle=\(\tilde{\chi}_1^0\)],
        };
        \caption{Long-lived gluino decay through highly virtual squark.}
        \label{displaced_process}
        \end{center} \end{figure}

@@

        The authors minimize the effect of hadronic interactions by vetoing the material-dense regions of the detector (see Fig.~\ref{material_map}) and estimate the residual background by extrapolating from the low-mass region, which is dominated by hadronic interactions. After a dedicated study, the background from merged vertices was found to be negligible. Finally, the background from crossing tracks is determined from a model that adds fake tracks\fxnote{check} to data vertices.

@@

        Exactly zero events are observed with an expected background of $\num{0.02} \pm \num{0.02}$ events. The result is then interpreted as an upper limit on the gluino mass and pair-production cross section as a function of gluino lifetime, shown in Fig.~\ref{displaced_limit}. For an LSP neutralino mass of \SI{100}{\giga\electronvolt}, the gluino mass limit is approximately \SI{2000}{\giga\electronvolt} for lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range.

@@

    \subsection{Disappearing Tracks}
        Certain models with nearly mass-degenerate LSP and NLSP have the potential to present a remarkable signature: charged particles that seem to disappear in the tracker. Specifically, models with winos as the lightest gauginos, such as many AMSB models, will have a chargino NLSP with an observable lifetime (\SI{0.2}{\nano\s} is a typical estimate that eventually decays to the neutralino LSP and, commonly, a pion. The nearly degenerate masses have two important phenomenological roles: (1) increase the chargino lifetime, and (2) ensure low-momentum decay products. The low-momentum pion is then unlikely to be reconstructed, which results in a disappearing track~\cite{atlas_disappearing}.

        Both ATLAS and CMS looked for disappearing tracks during Run I, but the tracker improvements installed in ATLAS before 2015 datataking and CMS before 2017 datataking should increase sensitivity to decays at shorter radii. The ATLAS search presented here uses \SI{36.1}{\femto\barn^{-1}} of \SI{13}{\tera\electronvolt} data and is sensitive to LLP lifetimes of approximately \SI{10}{\pico\s} to \SI{10}{\nano\s}.

@@

        Disappearing track candidates are selected with a looser track reconstruction algorithm that runs over hits that were not associated to tracks by the standard reconstruction algorithm. The candidates must also meet isolation, transverse momentum ($\pT$), and quality requirements such as number of pixel layer hits. Finally, the disappearing tracks are selected by requiring exactly zero associated hits in the subsequent tracking layer. Event selection requires one disappearing track candidate, no electron or muon candidates (to suppress $\ttbar$ and $\W/\Z + \mathrm{jets}$ processes), and the $\MET$ and jet requirements associated with either of the two production modes described above.

        Disappearing track candidates can be faked by random combinations of unrelated hits from nearby particles or by tracks whose pixel hits and next tracking layer hits are not properly associated, either due to a hadron suffering a hard scatter or a lepton radiating a hard photon. These backgrounds come primarily from $\ttbar$ and $\W/\Z + \mathrm{jets}$ processes, and estimates for each type of background are obtained from data.

@@

    \subsection{Stopped Particles}
        \fxnote{too redundant with overview?}
        With long enough lifetimes and sufficient mass, LLPs could come to rest in the detector before decaying. The eventual decay would then lead to energy deposits in the calorimeters or hits in the muon system that could be out of time with proton bunch crossings. A recent CMS analysis looks for such a signal in dimuon final states using \SI{39}{\femto\barn^{-1}} of \SI{13}{\tera\electronvolt} data~\cite{cms_stopped}.

@@

        After estimating the background by extrapolating data from the cosmic-enriched region to the signal region, the authors calculate the expected background for LLP lifetimes ranging from $10^{-7}$ to $10^6$ s. In all cases, the expected background is less then one event. Exactly zero events are observed in every lifetime scenario. As shown in Fig.~\ref{stopped_limits}, these results are then interpreted as upper limits on the signal production cross section and mass\fxnote{mass of what?}, which are the first limits for stopped particles that decay to muons at the LHC.

@@

        HSCPs will often be misidentified or missed entirely by standard reconstruction algorithms and analysis criteria, which tend to assume $\frac{v}{c} \approx \num{1}$ and $\lvert q \rvert = e$. Inverting these assumptions provides two handles for a dedicated HSCP search using \SI{2.5}{\femto\barn^{-1}} of \SI{13}{\tera\electronvolt} data collected with the CMS detector in 2015. The search is interpreted in the context of three benchmark models: (1) HSCPs that form R-hadrons that either change the sign of their electric charge or become neutral before reaching the MS, (2) lepton-like HSCPs (quasi-stable sleptons) that are produced either through decays of squarks or gluinos or directly if squarks and gluinos are kinematically inaccessible, and (3) long-lived lepton-like fermions with arbitrary electric charge. This discussion will focus on the first two models, as they bear more directly on naturalness and SUSY.

@@

        The authors estimate the signal-region background using an ABCD method in which events are binned in a 2-D plane with uncorrelated axes and the ratio of events in two control regions is used to extrapolate into the signal region from a third control region.

@@

    \noindent \begin{figure}[htbp] \begin{center}
    \includegraphics[width=\textwidth]{images/atlas_susy_summary}
    \caption{Summary of current experimental limits on SUSY models from ATLAS searches.}
    \label{atlas_susy_summary}
    \end{center} \end{figure}

@@

        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.32\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/cornering_effective_100}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.32\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/cornering_rpv_100}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.32\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/cornering_rpv_effective_100}
        \end{center} \end{subfigure}
            \caption{Comparing experimental limits to $\Delta < 10$ naturalness bounds for Effective (left), RPV (center), and Effective+RPV (right) SUSY models ~\cite{cornering}.}
        \label{cornering_limits}
        \end{center} \end{figure}
